Hypertension statistics for US adults: an open-source web application for analysis and visualization of US NHANES data

### 2022-11-05 paper/manuscript.Rmd
progressing validation + tutorials

abbrvs <- list(
  "CI"   = 'confidence interval',
  "BP"   = "blood pressure",
  "SBP"  = "systolic blood pressure",
  "DBP"  = "diastolic blood pressure",
  "SD"   = 'standard deviation',
  "IQR"  = 'interquartile range',
  "JNC7" = "Seventh Joint National Committee",
  "ACC"  = "American College of Cardiology",
  "AHA"  = "American Heart Association",
  "CVD"  = 'cardiovascular disease'
)

@@

Byron C. Jaeger, PhD,^1^ 
Ligong Chen, PhD,^2^
Adam P. Bress, PharmD, MS,^3,4^ 
Kathryn Foti,^2^
Shakia T. Hardy,^2^
Bharat Poudel, MSPH,^2^
Ashley Christenson, BS,^2^ 
Lisandro D. Colantonio, PhD,^2^
Paul Muntner, PhD,^2^

^1^Department of Biostatistics and Data Science, Wake Forest School of Medicine, Winston-Salem, NC. `r single_par`
^2^Department of Epidemiology, University of Alabama at Birmingham. `r single_par`
^3^Informatics, Decision-Enhancement, and Analytic Sciences (IDEAS) Center, Veterans Affairs, Salt Lake City Health Care System, Salt Lake City, UT. `r single_par`
^4^Department of Population Health Sciences, University of Utah School of Medicine, Salt Lake City, UT. `r single_par`

@@

**Background**: Data from the US National Health and Nutrition Examination Survey (NHANES) are freely available and can be analyzed to produce hypertension statistics for the non-institutionalized US population. However, the analysis of these data requires statistical programming knowledge.

**Methods**: We developed and validated a web-based application that provides hypertension statistics for US adults using 10 cycles of NHANES data, 1999-2000 through 2017-2020, without requiring users to conduct statistical programming. The application’s interface allows users to estimate crude and age-adjusted means, quantiles, and proportions. Population counts can also be estimated. To demonstrate the application’s capabilities, we estimated hypertension statistics for US adults.

**Results**: The estimated mean systolic blood pressure (BP) declined from 123 mmHg in 1999-2000 to 120 mmHg in 2009-2010 and increased to 123 mmHg in 2017-2020. The age-adjusted prevalence of hypertension (i.e., systolic BP ≥130 mmHg, diastolic BP ≥80 mmHg or self-reported antihypertensive medication use) was 47.9% in 1999-2000, 43.0% in 2009-2010, and 44.7% in 2017-2020. In 2017-2020, 115.3 million US adults had hypertension. The age-adjusted prevalence of controlled BP, defined according to the 2017 American College of Cardiology/American Heart Association BP guideline, among US adults with hypertension was 9.6% in 1999-2000, 24.6% in 2013-2014, and 21.7% in 2017-2020. After age-adjustment and among US adults who self-reported taking antihypertensive medication, 27.4%, 48.5%, and 42.8% had controlled BP in 1999-2000, 2013-2014, and 2017-2020, respectively. The application was validated by replicating statistics from three publications.

**Conclusions**: The application developed and validated in the current study is publicly available and produced valid, transparent, and reproducible results.

@@

We developed an open-source web-based application that provides nationally representative BP and hypertension statistics for non-institutionalized US adults using NHANES data without requiring users to conduct statistical programming. The goal of the application is to increase the use of NHANES data for research and policy decision making with a focus on ensuring results are valid, transparent, and reproducible. In this manuscript, we review the design, development and validation of the application, present BP and hypertension statistics for US adults, and describe online tutorials and resources that may help users engage with the application.

@@

NHANES data were collected through in-home interviews and study examinations conducted at mobile examination centers. The interview included questions about demographics, health behaviors, medical history, and medication use. During the interview, the labels of medications that participants reported taking in the preceding 30 days were recorded. During the study examination, height, weight and BP were measured and blood and spot urine samples were collected. Of relevance to the application, blood samples were used to measure total and high-density lipoprotein cholesterol, glycated hemoglobin and serum creatinine, and the urine sample was used to measure albumin and creatinine and to conduct a pregnancy test. The protocol for obtaining BP measurements is available online.[@ostchega_national_2003] In brief, systolic and diastolic BP (SBP and DBP, respectively) were measured three times by trained and certified physicians. BP was measured using a mercury sphygmomanometer from 1999-2000 through 2015-2016 and using an oscillometric device in 2017-2020. The mean SBP and DBP levels were computed over all available measurements for each participant. The oscillometric SBP and DBP values were calibrated to the mercury device.[@muntner2022blood; @ostchega2021comparing] Antihypertensive medication classes were defined using those listed in the 2017 ACC/AHA BP guideline.[@whelton20182017] For the current analysis, we defined hypertension, BP control, and resistant hypertension according to the 2017 ACC/AHA BP guideline.[@whelton20182017] The application also has these variables defined according to the JNC7 definitions.[@chobanian2003seventh] Chronic kidney disease (CKD) was defined by estimated glomerular filtration rate < 60 ml/min/1.73 m^2^ calculated using the 2021 serum creatinine-based equation or albumin-to-creatinine ratio ≥ 30 mg/g.[@inker_new_2021] A list of BP, hypertension, and antihypertensive medication variables is provided in **Table 1** with full definitions for all variables provided in **Table S1**.

@@

There were 107,622 NHANES participants in the 10 cycles from 1999-2000 to 2017-2020. We restricted the dataset to adults ≥ 18 years of age. This exclusion was applied because statistics for BP levels and the prevalence of hypertension in children and adolescents are markedly different than for adults.[@muntner2004trends] We further restricted the population to participants who completed the in-home interview and study examination, with one or more SBP and DBP measurement, and who had data on self-reported antihypertensive medication use. After these exclusions were applied, the population in this application included 56,035 participants (**Figure S1**).

@@

A full summary of this application’s features and associated tutorials are available online.[@rpack_cardioStatsUSA] Briefly, users can select NHANES cycles from 1999-2000 to 2017-2020 to be analyzed. Estimates are weighted to represent the non-institutionalized US population and users may incorporate age-adjustment through direct standardization. Users can restrict analyses to subsets of participants (e.g., participants who self-reported taking antihypertensive medication). When population count estimates are requested, survey weights are calibrated to account for missing information on SBP, DBP or self-reported antihypertensive medication use. Users may tabulate or visualize summaries, and can present results for the overall population or in subgroups. All figures and datasets created with the web application can be downloaded and saved. Following CDC recommendations, unreliable statistical estimates are automatically suppressed. To increase precision and reliability of estimates, contiguous NHANES cycles can be combined.[@nhanes_tutorial_weights]

@@

The web application was created using Shiny,[@rpack_shiny; @rpack_shinyWidgets; @rpack_shinyalert] an open-source software package that translates code from the R programming language into HTML, CSS, or JavaScript commands that create a website interface.[@r_language] We created the “cardioStatsUSA” R package to provide additional details on the web application’s design and comprehensive documentation of its components.[@rpack_cardioStatsUSA] We validated the web application by using it to reproduce statistics reported in two prior studies and one CDC report.[@muntner2020htn; @muntner2022blood; @fryar_hypertension_2017]

@@

We performed statistical analyses to demonstrate core features of the application. We estimated the mean SBP for US adults by NHANES cycle, 1999-2000 through 2017-2020, with points showing the estimated means and error bars showing the 95% confidence interval (CI) for the estimate. We made bar charts presenting the age-adjusted prevalence of hypertension and the estimated number of US adults with hypertension. For age adjustment, we used the estimated age distribution of US adults from 1999 to 2020 as the standard (49.3%, 33.6%, 10.1% and 7.0% being 18 to 44, 45 to 64, 65 to 74 and ≥ 75 years of age, respectively). We created a table of the estimated race/ethnicity distribution of US adults with and without hypertension, separately. We demonstrated stratification by estimating the prevalence of hypertension by NHANES cycle for US adults with and without CKD. We showed the application’s ability to suppress output when statistical estimates are unstable by attempting to estimate the distribution of BP categories (SBP/DBP < 120/80 mm Hg, 120-129/<80 mm Hg, 130-139/80-89 mm Hg, 140-159/90-99 mm Hg and ≥ 160/100 mm Hg) among pregnant women in 2017-2020. We then showed that reliable estimates can be obtained for the distribution of BP categories among pregnant women by pooling NHANES cycles from 2009-2010 through 2017-2020.

@@

The age-adjusted prevalence of BP control among non-pregnant US adults with hypertension was lowest in 1999-2000 (9.6%), highest in 2013-2014 (24.5%), and 21.7% in 2017-2020 (**Figure 4; panel A**). Among US adults who self-reported taking antihypertensive medication, 27.6%, 48.4%, and 42.8% had controlled BP in 1999-2000, 2013-2014, and 2017-2020, respectively (**Figure 4; panel B**). The age-adjusted prevalence of resistant hypertension among non-pregnant US adults taking antihypertensive medication was lowest in 1999-2000 (14.3%), highest in 2005-2006 (21.4), and 16.6% in 2017-2020 (**Figure S4; panel A**). Among non-pregnant US adults who were taking three or more classes of antihypertensive medication, the prevalence of resistant hypertension was lowest in 2009-2010 (64.1%), highest in 2003-2004 (78.3%), and 67.9% in 2017-2020 (**Figure S4; panel B**).

@@

In the current manuscript, we present a web application that allows the calculation of nationally representative estimates for BP, hypertension and antihypertensive medication-related outcomes using publicly available NHANES data. Using this application, we generated crude and age-adjusted BP and hypertension statistics. Also, we generated statistics stratified by characteristics of US adults and pooling multiple NHANES cycles to obtain more precise estimates when working with the small sub-group of pregnant women. Following its validation, version 0.0.1 of the application was released and deployed in a publicly available server on **DATE TBD**.[@app_cardioStatsUSA] Researchers, clinicians, and the public can use the application to generate customized BP and hypertension statistics for US adults.

@@

NHANES was designed to obtain nationally representative estimates of the health and nutrition status of non-institutionalized US adults and is an ideal data source to obtain statistics related to hypertension. SBP and DBP were measured following a standardized protocol by trained and certified staff.  NHANES data are publicly available to download but working with these data requires understanding variable definitions and advanced programming and statistical knowledge. Challenges that users encounter when analyzing with NHANES include the need to download and merge multiple data files, even for a single NHANES cycle; combining multiple variables to create outcome definitions, which may require dealing with missing data and questionnaire skip patterns; harmonizing definitions across multiple NHANES cycles; and the analyses of complex survey design data. The web application that we present in the current manuscript was designed to reduce barriers to using NHANES so that these data can more fully inform public health decisions and future research related to hypertension.

@@

Several design decisions have been incorporated into the application. We required participants to have a single SBP and DBP to be included while some prior analyses required three SBP and DBP measurements.  We chose an approach that is consistent with several analyses conducted by CDC investigators.[@yoon_hypertension_2015; @fryar_hypertension_2017] However, mean BP and the prevalence of high BP may be lower if we required multiple BP measurements.[@handler_impact_2012] Any bias resulting from this decision is likely to be small as over 95% of adult NHANES participants with at least one SBP and DBP measurement had three SBP and DBP measurements. The application re-calibrates the NHANES weights for the estimation of population counts.[@coresh_prevalence_2003] This was done because participants missing data on SBP, DBP or antihypertensive medication use cannot have BP or hypertension-related outcomes. Weights were not re-calibrated when estimating proportions as participants missing data are removed from the numerator and denominator. Medication classes were coded using generic names and the drug classes in the 2017 ACC/AHA BP guideline. We recognize the NCHS recommends using Lexicon Plus®, a proprietary database, to categorize medication classes.[@nhanes_rxq_drug] While the categorization of most medications is identical using generic drug names and Lexicon Plus®, some differences exist. Many additional decisions were made regarding the definitions of variables, inclusion of study participants, and analytic approach. We sought to make decisions that would be widely acceptable and transparent to ensure the results could be described accurately.

@@

We were able to replicate several prior manuscripts using the application. However, we were not able to replicate some results. For example, a prior manuscript reported the prevalence of resistant hypertension defined by the 2017 ACC/AHA BP guideline to be 19.7% in 2009-2014.[@carey_prevalence_2019] When estimated by the application, the prevalence of resistant hypertension over this time period was 17.1%. The difference in the prevalence estimates can be attributed to the approach used to categorize medication into classes. The prior publication used Lexicon Plus®, which counted spironolactone as two drug classes, a potassium-sparing diuretic and an aldosterone antagonist. We included spironolactone as a single drug class, an aldosterone antagonist. Additionally, Lexicon Plus® includes Sotalol as a beta blocker and nitroglycerine as a direct vasodilator and we did not include these drugs as antihypertensive medications as they are not listed in the 2017 ACC/AHA BP guideline.[@whelton20182017] The differences in results between the application and this previously published manuscript emphasize that it may not be possible to always generate results that are identical to prior publications. Users should be aware of the choices made in defining variables as this may affect the statistical estimates generated using the application.

@@

The application has a number of features that can be used in combination to create highly customized statistics. Variables can be analyzed as outcomes or used to stratify results or restrict analyses to subsets of participants.  Combining the variables, tens of thousands of unique statistics can be estimated. In addition, users can generate statistics for US adults over 22 years, pooling results or stratifying the data. All results from the application can be saved as an image and included in scientific proposals or presentations. In addition, the ability to download results as a data set from the application allows users to further customize their results in tabular or graphical formats. The web application can also be used for educational purposes, as teachers can use it to analyze health outcomes for US adults interactively with students.

With an open-source software license and publicly available documentation and data, the web application we developed can be extended. The current manuscript described a set of hypertension variables and a sub-population of NHANES participants with data on BP and antihypertensive medication use. The application’s interface and its supporting R package allows additional variables to be added. In addition, the application can be extended to include data from additional NHANES cycles as they become available.

@@

This study has a number of strengths. We used NHANES data, which are publicly available, rigorously collected, and allow for estimation of nationally representative statistics. Also, we leveraged open-source software to ensure that our application is transparent, able to be extended, and freely available. This study also has several limitations. NHANES participants had their mean BP measured during a single visit, and BP guidelines recommend obtaining mean BP using at least two BP values measured on separate days.[@whelton20182017] The response rate for NHANES has declined from 1999-2000 through 2017-2020. Effects of this decline are unclear.

@@

# ACKNOWLEDGMENTS

**this section is under construction**

The views expressed in this paper are those of the authors and do not represent the official position of the National Institutes of Health (NIH), the National Heart, Lung, and Blood Institute, or the U.S. Government.

@@

### Table 1: Blood pressure, hypertension and antihypertensive medication variables that may be analyzed using the web application.

@@

ft_data <- nhanes_key %>% 
 .[module == 'htn'] %>% 
 .[class %in% classes] %>% 
 .[, class := factor(class, levels = classes)] %>% 
 .[order(class)] %>% 
 .[variable %in% bp_med_classes,
   class := "Antihypertensive medication classes"] %>% 
 select(class, variable, label) %>% 
 as_grouped_data(groups = 'class')

indent_meds_spanner <- 
 which(ft_data$class == "Antihypertensive medication classes")
indent_meds <- 
 which(ft_data$variable %in% bp_med_classes)

@@

@@

p1 <- nhanes_visualize(data = nhanes_data,
                       key = nhanes_key,
                       outcome_variable = 'bp_sys_mean',
                       time_values = 'all', 
                       geom = 'points',
                       size_point = 25,
                       size_error = 1)

```
Dots represent mean systolic blood pressure. Vertical lines represent the 95% confidence interval.

@@

```
Age adjustment was performed through direct standardization, using the estimated age distribution of US adults from 1999 to 2020 as the standard (49.3%, 33.6%, 10.1% and 7.0% being 18 to 44, 45 to 64, 65 to 74 and ≥ 75 years of age, respectively)

@@

@@

A. Among non-pregnant US adults with hypertension
```{r, fig.height=3.75}

@@

```
Age adjustment was performed through direct standardization, using the estimated age distribution of US adults with hypertension from 1999 to 2020 as the standard (26.4%, 43.4%, 17.0% and 13.2% being 18 to 44 years, 45 to 64, 65 to 74, and ≥ 75 years of age, respectively)

@@

B. Among non-pregnant US adults who self-report taking antihypertensive medication 
```{r, fig.height=3.75}
fig_print(p2)
```
Age adjustment was performed through direct standardization, using the estimated age distribution of US adults with hypertension from 1999 to 2020 as the standard (26.4%, 43.4%, 17.0% and 13.2% being 18 to 44 years, 45 to 64, 65 to 74, and ≥ 75 years of age, respectively)

@@

### Table S1: All variables included in the web-based application.

@@

ft <- ft_data %>% 
 as_flextable(hide_grouplabel = TRUE,
              col_keys = c('class', 'label', 'description')) %>% 
 set_header_labels(label = "All variables included in module",
                   description = "Variable definition") %>% 
 theme_box() %>% 
 bg(i = ~ !is.na(class), bg = 'grey80') %>%
 bg(i = indent_meds_spanner, bg = 'white') %>%
 italic(i = ~ !is.na(class)) %>%
 padding(i = ~ is.na(class), j = 1, padding.left = 10) %>% 
 padding(i = indent_meds_spanner, j = 1, padding.left = 10) %>% 
 padding(i = indent_meds, j = 1, padding.left = 20) %>% 
 width(j = 1, width = 2) %>%
 width(j = 2, width = 7) %>% 
 add_footer_lines(
  values = abbrvs_write(
   abbrvs[c('BP', 'SBP', 'DBP', 'ACC', 'AHA', 'JNC7')]
  )
 )

>>>>

abbrvs <- list(
  "CI"     = 'confidence interval',
  "BP"     = "blood pressure",
  "SBP"    = "systolic blood pressure",
  "DBP"    = "diastolic blood pressure",
  "SD"     = 'standard deviation',
  "IQR"    = 'interquartile range',
  "JNC7"   = "Seventh Joint National Committee",
  "ACC"    = "American College of Cardiology",
  "AHA"    = "American Heart Association",
  "ASCVD"  = "atherosclerotic cardiovascular disease",
  "CVD"    = 'cardiovascular disease',
  "HbA1c"  = "hemoglobin A1C",
  "ACE"    = "angiotensin-converting enzyme",
  "NHANES" = "National Health and Nutrition Examination Survey"
)

@@

Byron C. Jaeger, PhD,^1^ 
Ligong Chen, PhD,^2^
Kathryn Foti,^2^
Shakia T. Hardy,^2^
Adam P. Bress, PharmD, MS,^3,4^ 
Sean P. Kane, PharmD, BCPS,^5^
Jennifer S. Herrick, MS,^3,6^
Catherine Derington, PharmD, MS,^7^ 
Bharat Poudel, MSPH,^2^
Ashley Christenson, BS,^2^ 
Lisandro D. Colantonio, PhD,^2^
Paul Muntner, PhD,^2^

**AUTHOR ORDER NOT FINALIZED**

^1^Department of Biostatistics and Data Science, Wake Forest School of Medicine, Winston-Salem, NC. `r single_par`
^2^Department of Epidemiology, University of Alabama at Birmingham. `r single_par`
^3^Informatics, Decision-Enhancement, and Analytic Sciences (IDEAS) Center, Veterans Affairs, Salt Lake City Health Care System, Salt Lake City, UT. `r single_par`
^4^Department of Population Health Sciences, University of Utah School of Medicine, Salt Lake City, UT. `r single_par`
^5^Department of Pharmacy Practice, Rosalind Franklin University of Medicine and Science, North Chicago, IL, USA
^6^Department of Internal Medicine, University of Utah School of Medicine, Salt Lake City, UT
^7^ University of Colorado Skaggs School of Pharmacy and Pharmaceutical Sciences, Aurora, CO, USA
\newpage

@@

**Background**: Data from the US National Health and Nutrition Examination Survey (NHANES) are freely available and can be analyzed to produce hypertension statistics for the non-institutionalized US population. The analysis of these data requires statistical programming knowledge.

**Methods**: We developed and validated a web-based application that provides hypertension statistics for US adults using 10 cycles of NHANES data, 1999-2000 through 2017-2020, without requiring users to conduct statistical programming. The application’s interface allows users to estimate crude and age-adjusted means, quantiles, and proportions. Population counts can also be estimated. To demonstrate the application’s capabilities, we estimated hypertension statistics for US adults.The application was validated by replicating statistics from three publications.

**Results**: The estimated mean systolic blood pressure (BP) declined from 123 mmHg in 1999-2000 to 120 mmHg in 2009-2010 and increased to 123 mmHg in 2017-2020. The age-adjusted prevalence of hypertension (i.e., systolic BP ≥130 mmHg, diastolic BP ≥80 mmHg or self-reported antihypertensive medication use) was 47.9% in 1999-2000, 43.0% in 2009-2010, and 44.7% in 2017-2020. In 2017-2020, 115.3 million US adults had hypertension. The age-adjusted prevalence of controlled BP, defined by the 2017 American College of Cardiology/American Heart Association BP guideline, among US adults with hypertension was 9.7% in 1999-2000, 25.0% in 2013-2014, and 21.9% in 2017-2020. After age-adjustment and among US adults who self-reported taking antihypertensive medication, 27.6%, 48.5%, and 43.0% had controlled BP in 1999-2000, 2013-2014, and 2017-2020, respectively.

**Conclusions**: The application developed in the current study is publicly available and produced valid, transparent, and reproducible results.

@@

We developed an open-source web-based application that provides nationally representative BP and hypertension statistics for non-institutionalized US adults using NHANES data without requiring users to conduct statistical programming. The goal of the application is to increase the use of NHANES data for research and policy decision making with a focus on ensuring results are valid, transparent, and reproducible. In this manuscript, we review the design, development and validation of the application and present BP and hypertension statistics for US adults that were generated using the application.

@@

NHANES data were collected through an in-home interview and study examination conducted at a mobile examination center. The interview included questions about demographics, health behaviors, medical history, and medication use. During the interview, the labels of medications that participants reported taking in the preceding 30 days were recorded. Antihypertensive medication classes were defined using those listed in the 2017 ACC/AHA BP guideline.[@whelton20182017] During the study examination, height, weight and BP were measured and blood and spot urine samples were collected. Of relevance to the application, blood samples were used to measure total and high-density lipoprotein cholesterol, glycated hemoglobin and serum creatinine, and the urine sample was used to measure albumin and creatinine and to conduct a pregnancy test. The protocol for measuring BP is available online.[@ostchega_national_2003] In brief, systolic and diastolic BP (SBP and DBP, respectively) were measured three times by trained and certified physicians. BP was measured using a mercury sphygmomanometer from 1999-2000 through 2015-2016 and using an oscillometric device in 2017-2020. The mean SBP and DBP levels were computed over all available measurements for each participant. The oscillometric SBP and DBP values were calibrated to the mercury device.[@muntner2022blood; @ostchega2021comparing] For the current analysis, we defined hypertension, BP control, and resistant hypertension according to the 2017 ACC/AHA BP guideline.[@whelton20182017] The application also has these variables defined according to the JNC7 definitions.[@chobanian2003seventh] Chronic kidney disease (CKD) was defined by estimated glomerular filtration rate < 60 ml/min/1.73 m^2^ calculated using the 2021 serum creatinine-based equation or albumin-to-creatinine ratio > 30 mg/g.[@inker_new_2021] A list of BP, hypertension, and antihypertensive medication variables is provided in **Table 1** with full definitions for all variables provided in **Table S1**.

@@

```{r, echo=FALSE}
n_included <- 
 table.glue::table_value(sum(nhanes_data$svy_subpop_htn))
```

There were 107,622 NHANES participants in the 10 cycles from 1999-2000 to 2017-2020. We restricted the dataset to adults ≥ 18 years of age. This exclusion was applied because statistics for BP levels and the prevalence of hypertension in children and adolescents are markedly different than for adults.[@muntner2004trends] We further restricted the population to participants who completed the in-home interview and study examination, with one or more SBP and DBP measurement, and who had data on self-reported antihypertensive medication use. After these exclusions were applied, the population in this application included `r n_included` participants (**Figure S1**).

@@

A full summary of this application’s features and associated tutorials are available online.[@rpack_cardioStatsUSA] Briefly, users can select NHANES cycles from 1999-2000 to 2017-2020 to be analyzed. Estimates are weighted to represent the non-institutionalized US population and users may incorporate age-adjustment through direct standardization. Users can restrict analyses to subsets of participants (e.g., participants who self-reported taking antihypertensive medication). When population count estimates are requested, survey weights are calibrated to account for missing information on SBP, DBP or self-reported antihypertensive medication use. The results can be presented in tables or figures and for the overall population or in subgroups. All tables and figures created with the web application can be downloaded and saved. Following CDC recommendations, unreliable statistical estimates are automatically suppressed.[@parker2017national] To increase precision and reliability of estimates, contiguous NHANES cycles can be combined.[@nhanes_tutorial_weights]

@@

The web application was created using Shiny,[@rpack_shiny; @rpack_shinyWidgets; @rpack_shinyalert] an open-source software package that translates code from the R programming language into HTML, CSS, or JavaScript commands and creates a website interface.[@r_language] We created the “cardioStatsUSA” R package to provide additional details on the web application’s design and comprehensive documentation of its components.[@rpack_cardioStatsUSA] We validated the web application by using it to reproduce statistics reported in two prior studies and one CDC report.[@muntner2020htn; @muntner2022blood; @fryar_hypertension_2017]

@@

We performed statistical analyses to demonstrate core features of the application. We estimated the mean SBP for US adults by NHANES cycle, 1999-2000 through 2017-2020, with points and error bars representing the estimated means and 95% confidence intervals (CI), respectively. We made bar charts presenting the age-adjusted prevalence of hypertension and the estimated number of US adults with hypertension. For age adjustment, we used the estimated age distribution of US adults from 1999 to 2020 as the standard (49.3%, 33.6%, 10.1% and 7.0% being 18 to 44, 45 to 64, 65 to 74 and ≥ 75 years of age, respectively). We created a table of the estimated race/ethnicity distribution of US adults with and without hypertension, separately. We demonstrated stratification by estimating the prevalence of hypertension by NHANES cycle for US adults with and without CKD. We showed the application’s ability to suppress output when statistical estimates are unstable by attempting to estimate the distribution of BP categories (SBP/DBP < 120/80 mm Hg, 120-129/<80 mm Hg, 130-139/80-89 mm Hg, 140-159/90-99 mm Hg and ≥ 160/100 mm Hg) among pregnant women in 2017-2020. We then showed that reliable estimates can be obtained for the distribution of BP categories among pregnant women by pooling NHANES cycles from 2009-2010 through 2017-2020.

@@

The age-adjusted prevalence of BP control among non-pregnant US adults with hypertension was lowest in 1999-2000 (9.7%), highest in 2013-2014 (25.0%), and 21.9% in 2017-2020 (**Figure 4; panel A**). Among US adults with hypertension who self-reported taking antihypertensive medication, 27.6%, 48.5%, and 43.0% had controlled BP in 1999-2000, 2013-2014, and 2017-2020, respectively (**Figure 4; panel B**). The age-adjusted prevalence of resistant hypertension among non-pregnant US adults with hypertension, who self-reported taking antihypertensive medication and had ≥1 antihypertensive medication class identified during the pill-bottle review was lowest in 1999-2000 (14.3%), highest in 2005-2006 (21.4%), and 16.6% in 2017-2020 (**Figure S4; panel A**). Further restricting this analysis to those who were taking three or more classes of antihypertensive medication, the prevalence of resistant hypertension was lowest in 2009-2010 (64.1%), highest in 2003-2004 (78.3%), and 67.9% in 2017-2020 (**Figure S4; panel B**).

@@

In the current manuscript, we present a web application that allows the calculation of nationally representative estimates for BP, hypertension and antihypertensive medication-related outcomes using publicly available NHANES data. Using this application, we generated crude and age-adjusted BP and hypertension statistics. Also, we generated statistics stratified by characteristics of US adults and pooling multiple NHANES cycles to obtain more precise estimates when working with the small sub-group of pregnant women. Following its validation, version 0.0.1 of the application was released and deployed on a publicly available server on **DATE TBD (it's almost ready)**.[@app_cardioStatsUSA] Researchers, clinicians, and the public can use the application to generate customized BP and hypertension statistics for US adults.

@@

NHANES was designed to obtain nationally representative estimates of the health and nutrition status of non-institutionalized US adults and is an ideal data source to obtain statistics related to hypertension. SBP and DBP were measured following a standardized protocol by trained and certified physicians.  NHANES data are publicly available to download but working with these data requires understanding variable definitions and advanced programming and statistical knowledge. Challenges that users encounter when analyzing NHANES data include the need to download and merge multiple data files, even for a single NHANES cycle; combining multiple variables to create outcome definitions, which may require dealing with missing data and questionnaire skip patterns; harmonizing variables across multiple NHANES cycles; and the analysis of complex survey design data. The web application that we present in the current manuscript addresses these challenges, allowing NHANES data to more fully inform public health decisions and future research related to hypertension.

@@

Several design decisions have been incorporated into the application. We required participants to have a single SBP and DBP to be included while some prior analyses required three SBP and DBP measurements.[@muntner_potential_2018]  We chose an approach that is consistent with several analyses conducted by CDC investigators.[@yoon_hypertension_2015; @fryar_hypertension_2017] However, mean BP and the prevalence of high BP may be lower if we required multiple BP measurements.[@handler_impact_2012] Any bias resulting from this decision is likely to be small as over 95% of adult NHANES participants with at least one SBP and DBP measurement had three SBP and DBP measurements. The application re-calibrates the NHANES weights for the estimation of population counts.[@coresh_prevalence_2003] This was done because participants missing data on SBP, DBP or antihypertensive medication use cannot have BP or hypertension-related outcomes. Weights were not re-calibrated when estimating proportions as participants missing data are removed from the numerator and denominator. Medication classes were coded using generic names and the drug classes in the 2017 ACC/AHA BP guideline. We recognize the NCHS recommends using Lexicon Plus®, a proprietary database, to categorize medication classes.[@nhanes_rxq_drug] While the categorization of most medications is identical using generic drug names and Lexicon Plus®, some differences exist. Many additional decisions were made regarding the definitions of variables, inclusion of study participants, and analytic approach. We sought to make decisions that would be widely acceptable and transparent to ensure the results could be described accurately.

@@

We were able to replicate several prior manuscripts using the application.[@muntner2020htn; @muntner2022blood; @fryar_hypertension_2017] However, we were not able to replicate some results.[@carey_prevalence_2019] For example, a prior manuscript reported the prevalence of resistant hypertension defined by the 2017 ACC/AHA BP guideline to be 19.7% in 2009-2014.[@carey_prevalence_2019] When estimated by the application, the prevalence of resistant hypertension over this time period was 17.1%. The difference in the prevalence estimates can be attributed to the approach used to categorize medication into classes. The prior publication used Lexicon Plus®, which counted spironolactone as two drug classes, a potassium-sparing diuretic and an aldosterone antagonist. We included spironolactone as a single drug class, an aldosterone antagonist. Additionally, Lexicon Plus® includes Sotalol as a beta blocker and nitroglycerine as a direct vasodilator and we did not include these drugs as antihypertensive medications as they are not listed in the 2017 ACC/AHA BP guideline.[@whelton20182017] The differences in results between the application and this previously published manuscript emphasize that it may not be possible to always generate results that are identical to prior publications. Users should be aware of the choices made in defining variables as this may affect the statistical estimates generated using the application.

@@

The application has a number of features that can be used in combination to create highly customized statistics. Variables can be analyzed as outcomes or used to stratify results or restrict analyses to subsets of participants.  Combining the variables, tens of thousands of unique statistics can be estimated using the application. In addition, users can generate statistics for US adults from 1999 to March 2020, pooling results or stratifying the data. All results from the application can be saved as an image and included in scientific proposals or presentations. In addition, the ability to download results as a data set from the application allows users to further customize their results in tabular or graphical formats.

@@

This study has a number of strengths. We used NHANES data, which are publicly available, rigorously collected, and allow for estimation of nationally representative statistics. Also, we leveraged open-source software to ensure that our application is transparent and freely available. This study also has several limitations. NHANES participants had their mean BP measured during a single visit, and BP guidelines recommend obtaining mean BP using at least two BP values measured on separate days.[@whelton20182017] The response rate for NHANES has declined from 1999-2000 through 2017-2020. Effects of this decline are unclear.

@@

@@

### Table 1: Blood pressure, hypertension and antihypertensive medication variables that are available in the web application.

@@

ft_data <- nhanes_key %>% 
 .[module == 'htn'] %>% 
 .[class %in% classes] %>% 
 .[, class := factor(class, 
                     levels = classes,
                     labels = paste(classes, 'domain'))] %>% 
 .[order(class)] %>% 
 .[variable %in% bp_med_classes,
   class := "Antihypertensive medication classes"] %>% 
 select(class, variable, label) %>% 
 as_grouped_data(groups = 'class')

@@

```{r footer_functions}

footer_hover <- function(parenthetical,
                         hover_over){

 glue::glue(
 "The graph is identical to the web application's output. \\
 All values ({parenthetical}) can be obtained in the app \\
 by hovering over {hover_over}."
 )

@@

nhanes_visualize(data = nhanes_data,
                 key = nhanes_key,
                 outcome_variable = 'bp_sys_mean',
                 time_values = 'all', 
                 geom = 'points',
                 size_point = 25,
                 size_error = 1) %>% 
 fig_print()

```
Dots represent mean systolic blood pressure. Vertical lines represent the 95% confidence interval. `r single_par`

`r footer_hover(parenthetical = "estimated mean and the upper and lower limits of the 95% confidence interval", hover_over = "the mean systolic blood pressure")` `r single_par`

@@

```
Age adjustment was performed through direct standardization, using the estimated age distribution of US adults from 1999 to 2020 as the standard (49.3%, 33.6%, 10.1% and 7.0% being 18 to 44, 45 to 64, 65 to 74 and ≥ 75 years of age, respectively). `r single_par`

`r footer_hover(parenthetical = "estimated prevalence and the upper and lower limits of the 95% confidence interval", hover_over = "the bars")` `r single_par`

@@

```
`r footer_hover(parenthetical = "estimated count and the upper and lower limits of the 95% confidence interval", hover_over = "the bars")` `r single_par`

@@

A. Among non-pregnant US adults with hypertension `r single_par`
```{r, fig.height=3.75}

@@

```
Age adjustment was performed through direct standardization, using the estimated age distribution of US adults with hypertension from 1999 to 2020 as the standard (26.4%, 43.4%, 17.0% and 13.2% being 18 to 44 years, 45 to 64, 65 to 74, and ≥ 75 years of age, respectively) `r single_par`

`r footer_hover(parenthetical = "estimated prevalence and the upper and lower limits of the 95% confidence interval", hover_over = "the bars")` `r single_par`

Survey participants with missing values for pregnancy status were assumed to be non-pregnant `r single_par`

@@

B. Among non-pregnant US adults who self-report taking antihypertensive medication `r single_par`
```{r, fig.height=3.75}
fig_print(p2)
```
Age adjustment was performed through direct standardization, using the estimated age distribution of US adults with hypertension from 1999 to 2020 as the standard (26.4%, 43.4%, 17.0% and 13.2% being 18 to 44 years, 45 to 64, 65 to 74, and ≥ 75 years of age, respectively) `r single_par`

`r footer_hover(parenthetical = "estimated prevalence and the upper and lower limits of the 95% confidence interval", hover_over = "the bars")` `r single_par`

Survey participants with missing values for pregnancy status were assumed to be non-pregnant `r single_par`

@@

### Table S1: Definitions of variables included in the web-based application.

@@

ft <- ft_data %>% 
 as_flextable(hide_grouplabel = TRUE,
              col_keys = c('class', 'label', 'description')) %>% 
 set_header_labels(label = "All variables included in module",
                   description = "Variable definition") %>% 
 theme_box() %>% 
 bg(i = ~ !is.na(class), bg = 'grey80') %>%
 bg(i = indent_meds_spanner, bg = 'white') %>%
 italic(i = ~ !is.na(class)) %>%
 padding(i = ~ is.na(class), j = 1, padding.left = 10) %>% 
 padding(i = indent_meds_spanner, j = 1, padding.left = 10) %>% 
 padding(i = indent_meds, j = 1, padding.left = 20) %>% 
 width(j = 1, width = 2) %>%
 width(j = 2, width = 7) %>% 
 add_footer_lines(
  values = abbrvs_write(
   abbrvs[c("BP",
         "SBP",
         "DBP",
         "JNC7",
         "ACC",
         "AHA",
         "ASCVD",
         "CVD",
         "HbA1c",
         "ACE",
         "NHANES")]
  )
 )
